Appendix 3: Estimating species interactions

### 2015-10-26 Harris_2015/manuscript-materials/draft.tex
moar stuff

\documentclass[12pt,]{article}
\usepackage[compact]{titlesec}
\usepackage{fancyhdr}
\usepackage{hanging}
\pagestyle{fancy}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{setspace}
\usepackage{lineno}
\linenumbers
\usepackage{setspace}
\usepackage{geometry}
\usepackage{ifthen}
\geometry{verbose,letterpaper,tmargin=2.54cm, bmargin=2.54cm,lmargin=2.54cm,rmargin=2.54cm}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{4pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}

@@

\textbf{Abstract:} Estimating species interactions from co-occurrence
data is one of the most controversial tasks in community ecology. One
difficulty is that a single pairwise interaction can ripple through an
ecological network and produce surprising indirect consequences. For
example, the negative correlation between two competing species can be
reversed in the presence of a third species that is capable of
outcompeting both of them. Here, I apply models from statistical
physics, called Markov networks or Markov random fields, that can
predict the direct and indirect consequences of any possible species
interaction matrix. Interactions in these models can also be estimated
from observed co-occurrence rates via maximum likelihood. Using
simulated landscapes with known pairwise interaction strengths, I
evaluated Markov networks and six existing approaches. The Markov
networks consistently outperformed other methods, correctly isolating
direct interactions between species pairs even when indirect
interactions largely overpowered them. Two computationally efficient
approximations, based on controlling for indirect effects with linear or
generalized linear models, also performed well. Indirect effects
reliably caused a common null modeling approach to produce incorrect
inferences, however.

@@

To the extent that nontrophic species interactions (such as competition)
affect community assembly, ecologists might expect to find signatures of
these interactions in species composition data (MacArthur 1958, Diamond
1975). Despite decades of work and several major controversies, however
(Lewin 1983, Strong et al. 1984, Connor et al. 2013), existing methods
for detecting competition's effects on community structure are
unreliable (Gotelli and Ulrich 2009). In particular, species' effects on
one another can become lost in the complex web of indirect effects. For
example, the competitive interaction between the two shrub species in
Figure 1A can become obscured by their shared tendency to occur in
unshaded areas (Figure 1B). While ecologists have long known that
indirect effects can overwhelm direct ones at the landscape level
(Dodson 1970, Levine 1976), the vast majority of our methods for drawing
inferences from observational data do not control for these effects
(e.g. Diamond 1975, Strong et al. 1984, Gotelli and Ulrich 2009, Veech
2013, Pollock et al. 2014). To the extent that indirect interactions
like those in Figure 1 are generally important (Dodson 1970), existing
methods will thus not generally provide much evidence regarding species'
direct effects on one another. The goal of this paper is to resolve this
long-standing problem.

@@

\paragraph{Markov networks.}\label{markov-networks.}

Markov networks provide a framework for translating back and forth
between the conditional (all-else-equal) relationships among species
(Figure 1C) and the kinds of species assemblages that these
relationships produce. Here, I show how a set of conditional
relationships can be used to determine how groups of species can
co-occur. Methods for estimating conditional relationships from data are
discussed in the next section.

@@

\(p(\vec{y}; \alpha, \beta) \propto exp(\sum_{i}\alpha_i y_i + \sum_{<ij>}\beta_{ij}y_i y_j),\)

@@

\noindent where the second sum is over all \(\binom{n}{2}\) species
pairs. In this model, \(\alpha_{i}\) is an intercept term determining
the amount that the presence of species \(i\) contributes to the
log-probability of \(\vec{y}\); it directly controls the prevalence of
species \(i\). Similarly, \(\beta_{ij}\) is the amount that the
co-occurrence of species \(i\) and species \(j\) contributes to the
log-probability; it controls the conditional relationship between two
species, i.e.~the probability that they will be found together, after
controlling for the other species in the network (Figure 2A, Figure 2B).
For example, \(\beta_{ij}\) might have a value of \(+2\) for two
mutualists, indicating that the odds of observing one species are
\(e^2\) times higher in sites where its partner is present than in
comparable sites where its partner is absent. Because the relative
probability of a presence-absence vector increases when
positively-associated species co-occur and decreases when
negatively-associated species co-occur, the model tends---all else
equal---to produce assemblages that have many pairs of
positively-associated species and relatively few pairs of
negatively-associated species (exactly as an ecologist might expect).

Of course, if all else is \emph{not} equal (e.g.~Figure 1, where the
presence of one competitor is associated with release from another
competitor), then species' overall co-occurrence patterns can be more
complicated. Determining the marginal relationships between species from
their conditional interactions entails summing over the different
possible assemblages (Figure 2B). This becomes intractable when the
number of possible assemblages is large, though several methods beyond
the scope of this paper can be employed to keep the calculations
feasible (Salakhutdinov 2008, Lee and Hastie 2012). Alternatively, as
noted below, some common linear and generalized linear methods can also
be used as computationally efficient approximations to the full network
(Lee and Hastie 2012, Loh and Wainwright 2013).

\paragraph{\texorpdfstring{Estimating \(\alpha\) and \(\beta\)
coefficients from presence-absence
data.}{Estimating \textbackslash{}alpha and \textbackslash{}beta coefficients from presence-absence data.}}\label{estimating-alpha-and-beta-coefficients-from-presence-absence-data.}

In the previous section, the values of \(\alpha\) and \(\beta\) were
known and the goal was to make predictions about possible species
assemblages. In practice, however, ecologists will often need to
estimate the parameters from an observed co-occurrence matrix (i.e.~from
a set of independent \(\vec{y}\) vectors indicating which species are
present at each site on the landscape). When the number of species is
reasonably small, one can compute exact maximum likelihood estimates for
all of the \(\alpha\) and \(\beta\) coefficients given a
presence-absence matrix by optimizing \(p(\vec{y}; \alpha, \beta)\).
Fully-observed Markov networks like the ones considered here have
unimodal likelihood surfaces (Murphy 2012), ensuring that this procedure
will always converge on the global maximum. This maximum represents the
unique combination of \(\alpha\) and \(\beta\) coefficients that would
be expected to produce exactly the observed co-occurrence frequencies on
average (i.e.~maximizing the likelihood ensures that the sufficient
statistics of the model distribution match the sufficient statistics of
the data; Murphy 2012). I used the rosalia package (Harris 2015a) for
the R programming language (R Core Team 2015) to optimize the Markov
network parameters. The package was named after Santa Rosalia, the
patron saint of biodiversity, whose mythos played an important
rhetorical role in the null model debates of the 1970's and 1980's
(Lewin 1983).

\paragraph{Simulated landscapes.}\label{simulated-landscapes.}

I simulated several sets of landscapes using known parameters as a test
bed for different statistical methods. The first set of landscapes
included the three competing species shown in Figure 1. For each of 1000
replicates, I generated a landscape with 100 sites by sampling from a
probability distribution defined by the figure's interaction
coefficients (Appendix 1). Each of the methods described below was then
evaluated on its ability to correctly infer that the two shrub species
competed with one another, despite their frequent co-occurrence.

I then generated landscapes with up to 20 interacting species at 25,
200, or 1600 sites using three increasingly complex models (50
replicates for each combination of size and model). As described in
Appendix {[}{[}X{]}{]}, I randomly drew the ``true'' coefficient values
for each replicate landscape in these simulations so that most species
pairs interacted negligibly, a few pairs interacted very strongly, and
competition was three times more common than facilitation.

The first set of 20-species landscapes, like the landscapes with three
species, were generated directly from a Markov network to ensure that
the model could recover the parameters used to generate the ``observed''
co-occurrence data. Then, I added two environmental factors that varied
from location to location across the simulated landscapes, and simulated
a new set of co-occurrence data so that species' \(\alpha\) coefficients
depended on these environmental variables. The latter set of simulated
landscapes provide an important test of the methods' ability to
distinguish co-occurrence patterns that were generated from pairwise
biotic interactions from those that were generated by external forces
like abiotic environmental filtering. This task was made especially
difficult because---as with most analyses of presence-absence data for
co-occurrence patterns---the inference procedure did not have access to
any information about the environmental or spatial variables that helped
shape the landscape (cf Connor et al. 2013). Finally, I simulated the
population dynamics of 20 interacting species, as described in Appendix
{[}{[}X{]}{]}

\paragraph{Recovering species interactions from simulated
data.}\label{recovering-species-interactions-from-simulated-data.}

I compared seven techniques for determining the sign and strength of the
associations between pairs of species from simulated data (Appendix 3).
First, I used the rosalia package (Harris 2015a) to fit Markov network
models, as described above. For the analyses with 20 species, I added a
weak regularizer (equivalent to a logistic prior with scale 2) to ensure
that the model estimates were always finite {[}{[}Appendix{]}{]}.

I also evaluated six alternative methods: five from the existing
literature, plus a novel combination of two of these methods. The first
alternative interaction metric was the sample correlation between
species' presence-absence vectors, which summarizes their marginal
association. Next, I used partial correlations, which summarize species'
conditional relationships. This approach is common in molecular biology
(Friedman et al. 2008), but less so in ecology (see Albrecht and Gotelli
2001 @faisal\_inferring\_2010 for two exceptions). In the context of
non-Gaussian data, the partial correlation can be thought of as a
computationally efficient approximation to the full Markov network model
(Loh and Wainwright 2013). Because partial correlations are undefined
for landscapes with perfectly-correlated species pairs, I used a
regularized estimate based on James-Stein shrinkage, as implemented in
the corpcor package's \texttt{pcor.shrink} function with the default
settings (Schäfer et al. 2014).

@@

\paragraph{Evaluating model
performance.}\label{evaluating-model-performance.}

For the simulated landscapes based on Figure 1, I assessed whether each
method's test statistic indicated a positive or negative relationship
between the two shrubs (Appendix 1). For the null model (Pairs), I
calculated statistical significance using its \(Z\)-score. For the
Markov network, I used the Hessian matrix to generate approximate
confidence intervals and noted whether these intervals included zero.

@@

\paragraph{Three species.}\label{three-species.}

As shown in Figure 1, the marginal relationship between the two shrub
species was positive---despite their competition for space at a
mechanistic level---due to indirect effects of the dominant tree
species. As a result, the correlation between these species was positive
in 94\% of replicates, and the randomization-based null model falsely
reported positive associations 100\% of the time. Worse, more than 98\%
of these false conclusions were statistically significant. The partial
correlation and Markov network estimates, on the other hand, each
correctly isolated the direct negative interaction between the shrubs
from their positive indirect interaction 94\% of the time (although the
confidence intervals overlapped zero in most replicates).

\paragraph{Twenty species.}\label{twenty-species.}

Despite some variability across contexts (Figure 3), the four methods
that controlled for indirect effects (the Markov network, the
generalized linear models, and the two partial correlation-based
methods) clearly outperformed those that did not. Of these, the Markov
network consistently performed the best, especially on the smaller
landscapes (which are also the ones that are most representative of
typical analyses in this field; Gotelli and Ulrich 2009).

Of the methods that did not control for indirect effects, Figure 3 shows
that simple correlation coefficients provided a more reliable indicator
of species' true interaction strengths than either the joint species
distribution model (BayesComm) or the null model (Pairs). Interestingly,
the estimates from these three approaches were tightly correlated
{[}{[}Figure 3B{]}{]}. This suggests that the effect sizes from the null
model (and, to a lesser extent, the correlation matrices from joint
species distribution models) only contain noisy versions of the same
information that could be obtained more easily and interpretably by
calculating correlation coefficients between species' presence-absence
vectors.

Bootstrap resampling indicated that the above ranking of the different
methods was robust (Appendix 3). In particular, the 95\% confidence
interval of the bootstrap distribution indicated that the Markov
network's overall \(R^2\) value was between {[}{[}14 and 18{]}{]}
percent higher than the second-most effective method (generalized linear
models) and between {[}{[}2.12 and 2.38{]}{]} times higher than could be
achieved by the null model (Pairs).

@@

Markov networks---particularly the Ising model for binary
networks---have been studied for nearly a century (Cipra 1987), and the
models' properties, capabilities, and limits are well-understood in a
huge range of applications. Using the same framework for species
interactions would thus allow ecologists to tap into an enormous set of
existing discoveries and techniques for dealing with indirect effects,
stability, and alternative stable states. Numerous other extensions are
possible: for example, the states of the interaction network can be
modeled as a function of the local abiotic environment (Lee and Hastie
2012), which would provide a rigorous and straightforward approach to
the difficult and important task of incorporating whole networks of
biotic interactions into species distribution models (Pollock et al.
2014), leading to a better understanding of the interplay between biotic
and abiotic effects on community structure. There are even methods
(Whittam and Siegel-Causey 1981, Tjelmeland and Besag 1998) that would
allow one species to affect the sign or strength of the relationship
between two other species, tipping the balance between facilitation and
exploitation (Bruno et al. 2003).

Finally, the results presented here have important implications for
ecologists' continued use of null models for studying species
interactions. Null and neutral models can be useful for clarifying our
thinking about the numerical consequences of species' richness and
abundance patterns (Harris et al. 2011, Xiao et al. 2015), but
deviations from a particular null model must be interpreted with care
(Roughgarden 1983). Even in small networks with three species, it may
simply not be possible to implicate individual species pairs or specific
ecological processes like competition by rejecting a general-purpose
null (Gotelli and Ulrich 2009), especially when the test statistic is
effectively just a correlation coefficient (Figure 3B). Simultaneous
estimation of multiple ecological parameters seems like a much more
promising approach: to the extent that the models' relative performance
on real data sets is similar to the range of results shown in Figure 3A,
scientists in this field could often double their explanatory power by
switching from null models to Markov networks (or increase it
substantially with linear or generalized linear approximations).
Regardless of the methods ecologists ultimately choose, controlling for
indirect effects could clearly improve our understanding of species'
direct effects on one another and on community structure.

\paragraph{Acknowledgements:}\label{acknowledgements}

This research was funded by a Graduate Research Fellowship from the US
National Science Foundation and benefited greatly from discussions with
A. Sih, M. L. Baskett, R. McElreath, R. J. Hijmans, A. C. Perry, and C.
S. Tysor. Additionally, A. K. Barner, E. Baldridge, E. P. White, D. Li,
D. L. Miller, N. Golding, N. J. Gotelli, C. F. Dormann, and two
anonymous reviewers provided very helpful feedback on the text.

@@

\textbf{Figure 1: A.} A small network of three competing species. The
tree (top) tends not to co-occur with either of the two shrub species,
as indicated by the strongly negative coefficient linking them. The two
shrub species also compete with one another, but more weakly (circled
coefficient). \textbf{B.} In spite of the competitive interactions
between the two shrub species, their shared tendency to occur in
locations without trees makes their occurrence vectors positively
correlated (circled). \textbf{C.} Controlling for the tree species'
presence with a conditional (all-else-equal) approach such as a partial
covariance or a Markov network leads to correct identification of the
negative shrub-shrub interaction (circled).

\textbf{Figure 2: A.} A small Markov network with two species, defined
by its \(\alpha\) and \(\beta\) values. The abiotic environment favors
the occurrence of each species (\(\alpha >0\)), particularly species 2
(\(\alpha_2 > \alpha_1\)). The negative \(\beta\) coefficient linking
these two species implies that they co-occur less than expected under
independence. \textbf{B.} These coefficients determine the probabilities
of all four possible presence-absence combinations for Species 1 and
Species 2. \(\alpha_1\) is added to the exponent whenever Species 1 is
present (\(y_1 = 1\)), but not when it is absent (\(y_1 = 0\)).
Similarly, the exponent includes \(\alpha_2\) only when species \(2\) is
present (\(y_2 = 1\)), and \(\beta\) only when both are present
(\(y_1y_2 = 1\)). The normalizing constant \(Z\), ensures that the four
probabilities sum to 1. In this case, \(Z\) is about 18.5. \textbf{C.}
The expected frequencies of all possible co-occurrence patterns between
the two species of interest, as calculated in the previous panel.
\textbf{D.} If, as assumed in null model simulations, \(\beta_{12}\)
equaled zero (e.g.~if the species no longer competed for the same
resources), then the reduction in competition would allow each species
to increase its overall occurrence rate.

\textbf{Figure 3: A.} Proportion of variance in interaction coefficients
explained by each method versus number of sampled locations. \textbf{B.}
The \(Z\)-scores produced by the null model (``Pairs'') for each pair of
species can be predicted using the correlation between the
presence-absence vectors of those same species and from the number of
sites on the landscape.

@@

Salakhutdinov, R. 2008. Learning and evaluating Boltzmann machines.
Technical Report UTML TR 2008-002, Department of Computer Science,
University of Toronto, Dept. of Computer Science, University of Toronto.

>>>>

\documentclass[12pt,]{article}
\usepackage[compact]{titlesec}
\usepackage{fancyhdr}
\usepackage{hanging}
\pagestyle{fancy}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{setspace}
\usepackage{lineno}
\linenumbers
\usepackage{setspace}
\usepackage{geometry}
\usepackage{ifthen}
\geometry{verbose,letterpaper,tmargin=2.35cm, bmargin=2.35cm,lmargin=2.44cm,rmargin=2.44cm}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{4pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}

@@

\textbf{Abstract:} Estimating species interactions from co-occurrence
data is one of the most controversial tasks in community ecology. One
difficulty is that a single pairwise interaction can ripple through an
ecological network and produce surprising indirect consequences. For
example, the negative correlation between two competing species can be
reversed in the presence of a third species that is capable of
outcompeting both of them. Here, I apply models from statistical
physics, called Markov networks or Markov random fields, that can
predict the direct and indirect consequences of any possible species
interaction matrix. Interactions in these models can also be estimated
from observed co-occurrence rates via maximum likelihood, controlling
for indirect effects. Using simulated landscapes with known pairwise
interaction strengths, I evaluated Markov networks and six existing
approaches. The Markov networks consistently outperformed other methods,
correctly isolating direct interactions between species pairs even when
indirect interactions or abiotic factors largely overpowered them. Two
computationally efficient approximations, based on controlling for
indirect effects with linear or generalized linear models, also
performed well. Indirect effects reliably caused a common null modeling
approach to produce incorrect inferences, however.

@@

To the extent that nontrophic species interactions (such as competition)
affect community assembly, ecologists might expect to find signatures of
these interactions in species composition data (MacArthur 1958, Diamond
1975). Despite decades of work and several major controversies, however
(Lewin 1983, Strong et al. 1984, Connor et al. 2013), existing methods
for detecting competition's effects on community structure are
unreliable (Gotelli and Ulrich 2009). In particular, species' effects on
one another can become lost in the complex web of indirect effects. For
example, the competitive interaction between the two shrub species in
Figure 1A can become obscured by their shared tendency to occur in
unshaded areas (Figure 1B). While ecologists have long known that
indirect effects can overwhelm direct ones at the landscape level
(Dodson 1970, Levine 1976), the vast majority of our methods for drawing
inferences from observational data do not control for these effects
(e.g. Diamond 1975, Strong et al. 1984, Gotelli and Ulrich 2009, Veech
2013, Pollock et al. 2014). To the extent that indirect interactions
like those in Figure 1 are generally important (Dodson 1970), existing
methods will thus not provide much evidence regarding species
interactions.

@@

\textbf{Markov networks.} Markov networks provide a framework for
translating back and forth between the conditional (all-else-equal)
relationships among species (Figure 1C) and the kinds of species
assemblages that these relationships produce. Here, I show how a set of
conditional relationships can be used to determine how groups of species
can co-occur. Methods for estimating conditional relationships from data
are discussed in the next section.

@@

\(\displaystyle{p(\vec{y}; \alpha, \beta) \propto exp(\sum_{i}\alpha_i y_i + \sum_{<ij>}\beta_{ij}y_i y_j),}\)

@@

\noindent where the second sum is over all \(\frac{1}{2}n(n-1)\) pairs
of \(n\) species. In this model, \(\alpha_{i}\) is an intercept term
determining the amount that the presence of species \(i\) contributes to
the log-probability of \(\vec{y}\); it directly controls the prevalence
of species \(i\). Similarly, \(\beta_{ij}\) is the amount that the
co-occurrence of species \(i\) and species \(j\) contributes to the
log-probability; it controls the conditional relationship between two
species, i.e.~the probability that they will be found together, after
controlling for the other species in the network (Figure 2A, Figure 2B).
For example, \(\beta_{ij}\) might have a value of \(+2\) for two
mutualists, indicating that the odds of observing one species are
\(e^2\) times higher in sites where its partner is present than in
comparable sites where its partner is absent.

Because the relative probability of a presence-absence vector increases
when positively-associated species co-occur and decreases when
negatively-associated species co-occur, the model tends---all else
equal---to produce assemblages that have many pairs of
positively-associated species and relatively few pairs of
negatively-associated species (exactly as an ecologist might expect). Of
course, if all else is \emph{not} equal (e.g.~Figure 1, where the
presence of one competitor is associated with release from another
competitor), then species' overall co-occurrence patterns can be more
complicated, and may require summing over the different possible
assemblages (Figure 2B) to determine the overall outcomes.

\noindent \textbf{Estimating $\alpha$ and $\beta$ coefficients from presence-absence data.}
In the previous section, the values of \(\alpha\) and \(\beta\) were
known and the goal was to make predictions about possible species
assemblages. In practice, however, ecologists will often need to
estimate the parameters from an observed co-occurrence matrix (i.e.~from
a set of independent \(\vec{y}\) vectors indicating which species are
present at each site on the landscape). When the number of species is
reasonably small, one can compute exact maximum likelihood estimates for
all of the \(\alpha\) and \(\beta\) coefficients given a
presence-absence matrix by optimizing \(p(\vec{y}; \alpha, \beta)\).
Fully-observed Markov networks like the ones considered here have
unimodal likelihood surfaces (Murphy 2012), ensuring that this procedure
will always converge on the global maximum. This maximum represents the
unique combination of \(\alpha\) and \(\beta\) coefficients that would
be expected to produce exactly the observed co-occurrence frequencies on
average (i.e., maximizing the likelihood ensures that the sufficient
statistics of the model distribution match the sufficient statistics of
the data; Murphy 2012). I used the rosalia package (Harris 2015a) for
the R programming language (R Core Team 2015) to optimize the Markov
network parameters.

\noindent \textbf{Simulated landscapes.} I simulated several sets of
landscapes using known parameters as a test bed for different
statistical methods. The first set of landscapes included the three
competing species shown in Figure 1. For each of 1000 replicates, I
generated a landscape with 100 sites by sampling from a probability
distribution defined by the figure's interaction coefficients (Appendix
1). Each of the methods described below was then evaluated on its
ability to correctly infer that the two shrub species competed with one
another, despite their frequent co-occurrence.

I then generated landscapes with up to 20 interacting species at 25,
200, or 1600 sites using three increasingly complex models (50
replicates for each combination of size and model). As described in
Appendix {[}{[}X{]}{]}, I randomly drew the ``true'' coefficient values
for each replicate landscape so that most species pairs interacted
negligibly, a few pairs interacted very strongly, and competition was
three times more common than facilitation.

The first set of 20-species landscapes, like the landscapes with three
species, were generated directly from a Markov network to ensure that
the model could recover the parameters used to generate the ``observed''
co-occurrence data. Then, I added two environmental factors that varied
from location to location across the simulated landscapes, and simulated
a new set of co-occurrence data so that species' \(\alpha\) coefficients
depended on the local environment. The latter set of simulated
landscapes provide an important test of the methods' ability to
distinguish co-occurrence patterns that were generated from pairwise
biotic interactions from those that were generated by external forces
like abiotic environmental filtering. This task was made especially
difficult because---as with most analyses of presence-absence data for
co-occurrence patterns---the inference procedure did not have access to
any information about the environmental or spatial variables that helped
shape the landscape (cf Connor et al. 2013). I generated the final set
of landscapes by simulating the population dynamics of 20 interacting
species, as described in Appendix {[}{[}X{]}{]}.

\noindent \textbf{Recovering species interactions from simulated data.}
I compared seven techniques for determining the sign and strength of the
associations between pairs of species from simulated data (Appendix 3).
First, I used the rosalia package (Harris 2015a) to fit Markov network
models, as described above. For the analyses with 20 species, I added a
weak regularizer (equivalent to a logistic prior with scale 2) to ensure
that the model estimates were always finite {[}{[}Appendix{]}{]}.

I also evaluated six alternative methods: five from the existing
literature, plus a novel combination of two of these methods. The first
alternative interaction metric was the sample correlation between
species' presence-absence vectors, which summarizes their marginal
association. Next, I used partial correlations, which summarize species'
conditional relationships. This approach is common in molecular biology
(Friedman et al. 2008), but is rare in ecology (see Albrecht and Gotelli
(2001) and Faisal et al. (2010) for two exceptions). In the context of
non-Gaussian data, the partial correlation can be thought of as a
computationally efficient approximation to the full Markov network model
(Loh and Wainwright 2013). Because partial correlations are undefined
for landscapes with perfectly-correlated species pairs, I used a
regularized estimate based on James-Stein shrinkage, as implemented in
the corpcor package's \texttt{pcor.shrink} function with the default
settings (Schäfer et al. 2014).

@@

\noindent \textbf{Evaluating model performance.} For the simulated
landscapes based on Figure 1, I assessed whether each method's test
statistic indicated a positive or negative relationship between the two
shrubs (Appendix 1). For the null model (Pairs), I calculated
statistical significance using its \(Z\)-score. For the Markov network,
I used the Hessian matrix to generate approximate confidence intervals
and noted whether these intervals included zero.

@@

\noindent \textbf{Three species.} As shown in Figure 1, the marginal
relationship between the two shrub species was positive---despite their
competition for space at a mechanistic level---due to indirect effects
of the dominant tree species. As a result, the correlation between these
species was positive in 94\% of replicates, and the randomization-based
null model falsely reported positive associations 100\% of the time.
Worse, more than 98\% of these false conclusions were statistically
significant. The partial correlation and Markov network estimates, on
the other hand, each correctly isolated the direct negative interaction
between the shrubs from their positive indirect interaction 94\% of the
time (although the confidence intervals overlapped zero in most
replicates).

\noindent
\textbf{Twenty species.} Despite some variability across contexts
(Figure 3), the four methods that controlled for indirect effects (the
Markov network, the generalized linear models, and the two partial
correlation-based methods) outperformed those that did not. Of these,
the Markov network consistently performed the best, especially on the
smaller landscapes (which are also the most representative of typical
analyses in this field; Gotelli and Ulrich 2009). As shown in Appendix
{[}{[}X{]}{]}, the standard errors associated with the estimates in
Figure 3 are all small (less than 0.01), so differences across methods
should not be attributed to sampling error.

Of the methods that did not control for indirect effects, Figure 3 shows
that simple correlation coefficients provided a more reliable indicator
of species' true interaction strengths than either the joint species
distribution model (BayesComm) or the null model (Pairs). The estimates
from these three approaches were tightly correlated (Figure 3B),
suggesting that the effect sizes from the null model (and, to a lesser
extent, the correlation matrices from joint species distribution models)
only contain noisy versions of the same information that could be
obtained more easily and interpretably by calculating correlation
coefficients between species' presence-absence vectors.

Whereas Figure 3 summarizes the accuracy of each model's point
estimates, Figure 4 compares the inferential value of the Markov network
versus the null model. When the ``true'' \(\beta\) coefficient is near
zero, the Markov network's 95\% confidence intervals usually include
zero (but not quite 95\% of the time), and it is rare for the model to
confidently attribute the wrong sign to a coefficient (Figure 4A). The
null model, by contrast, typically rejects the null hypothesis at the
\(p=0.05\) level regardless of the underlying parameters, and the sign
of its \(Z\)-scores is not a reliable indicator of the interaction's
actual sign.

@@

Markov networks---particularly the Ising model for binary
networks---have been studied for nearly a century (Cipra 1987), and the
models' properties, capabilities, and limits are well-understood. Using
the same framework for species interactions would thus allow ecologists
to tap into an enormous set of existing discoveries and techniques for
dealing with indirect effects, stability, and alternative stable states.
Numerous other extensions are possible: for example, the states of the
interaction network can be modeled as a function of the local abiotic
environment (Lee and Hastie 2012), which would provide a rigorous and
straightforward approach to the difficult and important task of
incorporating whole networks of biotic interactions into species
distribution models (Pollock et al. 2014), leading to a better
understanding of the interplay between biotic and abiotic effects on
community structure. Alternatively, models could allow one species to
affect the interactions between two other species (Tjelmeland and Besag
1998), tipping the balance between facilitation and exploitation (Bruno
et al. 2003).

Finally, the results presented here have important implications for
ecologists' continued use of null models for studying species
interactions. Null and neutral models can be useful for clarifying our
thinking about the numerical consequences of species' richness and
abundance patterns (Harris et al. 2011, Xiao et al. 2015), but
deviations from a particular null model must be interpreted with care
(Roughgarden 1983). Even in small networks with three species, it may
simply not be possible to implicate individual species pairs or specific
ecological processes like competition by rejecting a general-purpose
null (Gotelli and Ulrich 2009), especially when the test statistic is
effectively just a correlation coefficient {[}{[}Figure 3B{]}{]}. When a
non-null backdrop is not controlled for, Type I error rates can
skyrocket, the apparent sign of the interaction can change, and null
models can lead to reliably incorrect inferences (Figure 1, Figure 4).

Controlling for indirect effects via simultaneous estimation of multiple
ecological parameters seems like a much more promising approach: to the
extent that the models' relative performance on real data sets is
similar to the range of results shown in Figure 3A, scientists in this
field could more than double their explanatory power by switching from
null models to Markov networks (or increase it nearly as much with
linear or generalized linear approximations). Regardless of the methods
ecologists ultimately choose, controlling for indirect effects could
clearly improve our understanding of species' direct effects on one
another and on community structure.

\noindent \textbf{Acknowledgements:} This research was funded by a
Graduate Research Fellowship from the US National Science Foundation and
benefited greatly from discussions with A. Sih, M. L. Baskett, R.
McElreath, R. J. Hijmans, A. C. Perry, and C. S. Tysor. Additionally, A.
K. Barner, E. Baldridge, E. P. White, D. Li, D. L. Miller, N. Golding,
N. J. Gotelli, C. F. Dormann, and two anonymous reviewers provided very
helpful feedback on the text.

@@

\noindent \textbf{Figure 1: A.} A small network of three competing
species. The tree (top) tends not to co-occur with either of the two
shrub species, as indicated by the strongly negative coefficient linking
them. The two shrub species also compete with one another, but more
weakly (circled coefficient). \textbf{B.} In spite of the competitive
interactions between the two shrub species, their shared tendency to
occur in locations without trees makes their occurrence vectors
positively correlated (circled). \textbf{C.} Controlling for the tree
species' presence with a conditional (all-else-equal) approach such as a
partial covariance or a Markov network leads to correct identification
of the negative shrub-shrub interaction (circled).

\noindent \textbf{Figure 2: A.} A small Markov network with two species,
defined by its \(\alpha\) and \(\beta\) values. The abiotic environment
favors the occurrence of each species (\(\alpha >0\)), particularly
species 2 (\(\alpha_2 > \alpha_1\)). The negative \(\beta\) coefficient
linking these two species implies that they co-occur less than expected
under independence. \textbf{B.} These coefficients determine the
probabilities of all four possible presence-absence combinations for
Species 1 and Species 2. \(\alpha_1\) is added to the exponent whenever
Species 1 is present (\(y_1 = 1\)), but not when it is absent
(\(y_1 = 0\)). Similarly, the exponent includes \(\alpha_2\) only when
species \(2\) is present (\(y_2 = 1\)), and \(\beta\) only when both are
present (\(y_1y_2 = 1\)). The normalizing constant \(Z\), ensures that
the four probabilities sum to 1. In this case, \(Z\) is about 18.5.
\textbf{C.} The expected frequencies of all possible co-occurrence
patterns between the two species of interest, as calculated in the
previous panel. \textbf{D.} If, as assumed in null model simulations,
\(\beta_{12}\) equaled zero (e.g.~if the species no longer competed for
the same resources), then the reduction in competition would allow each
species to increase its overall occurrence rate.

\noindent \textbf{Figure 3: A.} Proportion of variance in interaction
coefficients explained by each method versus number of sampled
locations. \textbf{B.} The \(Z\)-scores produced by the null model
(``Pairs'') for each pair of species can be predicted using the
correlation between the presence-absence vectors of those same species
and from the number of sites on the landscape.

\noindent \textbf{Figure 4: A.} Inferences from the Markov network's
95\% confidence intervals on the ``no-environment'' simulated
landscapes, as a function of the ``true'' coefficient values.
\textbf{B.} Inferences from the \(Z\)-statistics estimated by the
``Pairs'' software for null modeling.

@@
